---
title: "Konza_Vegetation_Analysis"
author: "Emily Foster"
format: html
---

## Introduction 

This analysis will look at vegetation data from the Konza National Ecological   
Observatory Network (NEON) Site located in Manhattan KS. It will use    
tidyverse packages to wrangle and visualize data.    

## Loading the data

The data provided in csv files. We'll use the tidyverse package `readr` to load it into R. 

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")

# Read in Data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z (1).csv")


```

## data summarizing with `dplyr`

We will next use dplyr package to summarize some different aspects of the data.

```{r}
#| Label: summarize-with-dplyr

# first summary to make will be a count of number of observations in each family 

konza_data %>%
  group_by(family) %>%
  tally() %>%
  arrange(desc(n))
  
```

This analysis showed that Poaceae and Asteraceae are among the most common family   
found in tese plots. 

```{r}
#| label: mean-percent-cover-by-family

# the goal is to calculate the mean percent cover of each family 


```

