---
title: "Konza_Vegetation_Analysis"
author: "Emily Foster"
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Introduction 

This analysis will look at vegetation data from the Konza National Ecological   
Observatory Network (NEON) Site located in Manhattan KS. It will use    
tidyverse packages to wrangle and visualize data.    

## Loading the data

The data provided in csv files. We'll use the tidyverse package `readr` to load it into R. 

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")

# Read in Data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z (1).csv")


```

## data summarizing with `dplyr`

We will next use dplyr package to summarize some different aspects of the data.

```{r}
#| Label: summarize-with-dplyr

# first summary to make will be a count of number of observations in each family 

konza_data %>%
  group_by(family) %>%
  tally() %>%
  arrange(desc(n))
  
```

This analysis showed that Poaceae and Asteraceae are among the most common family found in these plots. 

```{r}
#| label: mean-percent-cover-by-family

# the goal is to calculate the mean percent cover of each family 

konza_data %>%
  group_by(family) %>%
  summarize(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover))

```

This analysis showed that the largest percent cover is the `NA` taxa, which means that it wasn't able to be identified, but the highest mean percent cover at 16% across plots was Fagaceae.

```{r}
#| Label: filter-by-elevation

konza_data %>%
  filter(elevation >390) %>%
  group_by(family) %>%
  summarize(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover))

```

Next we'll learn to use `mutate` which is a function to create a new column based on contents from an existing column 

```{r}
#| label: use-mutate

# select pulls out the column you want. 
# mutate creates a new column with whatever you want
# which can include manipulation of an exisiting column 
konza_data %>%
  select(scientificName, family, percentCover) %>%
  mutate(propCover = percentCover / 100)
```

Next up we'll use `ggplot` to make some figures.

```{r}
#| label: elevation-percent-by-family

konza_data %>%
  filter(!is.na(family)) %>%
  filter(family == "Asteraceae" | family == "Poaceae") %>%
  ggplot(aes( x = elevation, 
            y = percentCover,
            color = family)) +
  geom_point() + 
  geom_smooth() +
  labs(title = "Comparison of Asteraceae and Poaceae by elevation",
       y = "Percent Cover",
        X = "Elevation (m)")
  

```

## Homework figure 1

I have looked through the data and have decided to make a boxplot comparing the percent cover between the two plot types. I believe it may be interesting to see the distribution, which is what box plots are for. I will do this using the function `geom_boxplot`.

```{r}
#| label: percent-by-plot-type

konza_data %>% 
  filter(!is.na(family)) %>%
  ggplot(aes( x = plotType, 
              y = percentCover,
              color = family)) +
  geom_boxplot() +
  labs(title = "Comparison of Percent Cover by Plot Type",
        y = "Percent Cover",
        x = "Plot Type")
```

This plot shows that both Asteraceae and Poaceae has a wider distribution of plant cover in distributed plots compared to tower plots. Additionally, Fagaceae has a wider distribution of plant cover in tower plots when compared to distributed plots. 

## Homework Figure 2

I decided to investigate family occurrence within tower plots using a bar plot through `geom_bar`. I think a bar plot will be easier to view than a box plot.

```{r}
#| label: occurrence-by-tower-plot

konza_data %>%
  filter(!is.na(family)) %>%
  filter(plotType == "tower") %>%
  ggplot(aes( x = family)) +
  geom_bar() +
  labs(title = " Comparison of Family Occrance in Tower Plot Type",
       y = "Occurance",
       x = "Family")
  
```

This plot shows that Asteraceae, Poaceae and Fagaceae have higher occurrence in the Tower plot type. The family with the lowest occurrence within tower plots is Caprifoliaceae.

## Homework Figure 3

I decided to compare plot type and class with a plot using `geom_point`.

```{r}
#|label: class-by-plot-type

konza_data %>%
  filter(!is.na(family)) %>%
  ggplot(aes(x = plotType, 
             y = elevation,
             color = nlcdClass)) +
  geom_point() +
  labs(title = "Comparison of Plot Type and Class",
       y = "Elevation",
       x = " Plot Type")
  
```

This plot show that only grassland class plants are found in the tower plot, while both types are found in the distributed. Additionally, the grassland class plants are found at higher elevations than deciduous forest plants.
